---
title: "Single-cell RNA-seq Analysis of the Tumor and Immune Microenvironment in Multiple Myeloma CAR-T Patients"
subtitle: "Determinants of durable response, including baseline BCMA (TNFRSF17) expression"
author: "Bachisio Ziccheddu & Francesco Maura"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
# Global chunk options.
# eval = FALSE  -> the code is shown as a record of the analysis but is NOT
#                  re-run when the document is knitted. This lets the report
#                  build on any machine, without needing the raw data files
#                  or the original local file paths.
# Set eval = TRUE only if you want to actually execute everything end-to-end.
knitr::opts_chunk$set(
  echo    = TRUE,
  eval    = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

# Overview

This report documents the single-cell RNA-seq (scRNA-seq) analysis behind our study of
multiple myeloma patients treated with BCMA-targeted CAR-T therapy. Throughout, we ask
the same clinical question: **what distinguishes patients who achieve a durable response
(DR) from those who do not (NDR)?** — where a durable response is defined as
progression-free survival beyond 180 days.

To make the findings robust, every analysis combines **two independent patient cohorts**:

- **Moffitt** — our in-house dataset (`dataset = "moffitt"`)
- **Dhodapkar** — a public dataset (`dataset = "public"`)

The report is organized into **two parts**, each analyzing a different compartment of the
bone-marrow biopsy:

1. **Part 1 — Tumor cells.** The malignant plasma cells. We focus on baseline expression
   of *TNFRSF17* (the gene encoding **BCMA**, the CAR-T target), then broaden to
   genome-wide differential expression and pathway enrichment between DR and NDR patients.
2. **Part 2 — Immune microenvironment.** Every other, non-malignant cell type. We ask
   whether the composition and state of the immune cells at baseline — for example
   T-cell exhaustion — relate to response.

Both parts share the same backbone: load and quality-control each cohort, merge them,
remove the technical batch difference between datasets with **Harmony**, then compare
DR vs NDR — first at the single-cell level, and then by collapsing each patient into a
single "pseudobulk" profile for sample-level statistics.

---


# Part 1: Tumor cells

This part analyzes the **malignant plasma cells**. After quality control and integration
of the two cohorts, we isolate the plasma cells, correct for batch effects, and compare
durable versus non-durable responders — starting from the CAR-T target *TNFRSF17* (BCMA)
and expanding to genome-wide differential expression (DESeq2) and pathway enrichment
(GSEA).

---

## Libraries


```{r libraries, eval=FALSE}
library(SeuratObject)
library(Seurat)
library(harmony)
library(dplyr)
```

---

## Helper function: collapsing cells into "pseudobulk" per sample

Before the main analysis, we define a reusable function called `collapse_by_sample()`.

**What it does:** single-cell data has thousands of cells per
patient. For some comparisons we don't want thousands of cells — we want **one
expression profile per patient/sample**. This function takes all the cells belonging
to a sample and adds their gene counts together (a "pseudobulk" profile), into a per-cell average.

The function uses **sparse matrix operations**, which keeps it fast and memory-friendly
even on large objects. The `adjust` argument controls the output:

- `"none"` — raw summed counts per sample.
- `"percell_mean"` — average counts per cell (sum divided by the number of cells).
- `"percell_cpm"` — the per-cell average scaled to counts-per-million (handy for
  visualization, not recommended for differential expression).

```{r collapse-function, eval=FALSE}
# collapse_by_sample(): build pseudobulk matrices from a Seurat object
# - Uses sparse ops (fast, memory-safe)
# - adjust = "none" (sum counts), "percell_mean" (sum / #cells), or "percell_cpm"
#   * "percell_mean"  = average counts per cell in each sample
#   * "percell_cpm"   = CPM after per-cell averaging (useful for viz; not for DE)
#
collapse_by_sample <- function(
    seu,
    group.by = "sample_id",   # e.g., "orig.ident" or your sample column
    assay    = "RNA",
    slot     = "counts",
    adjust   = c("none","percell_mean","percell_cpm"),
    min_cells = 1,            # drop groups with < min_cells
    verbose  = TRUE
) {
  adjust <- match.arg(adjust)
  stopifnot(group.by %in% colnames(seu@meta.data))

  # pull sparse counts
  X <- Seurat::GetAssayData(seu, assay = assay, slot = slot)
  if (!inherits(X, "dgCMatrix")) {
    stop("Counts are not sparse (dgCMatrix). Convert or ensure slot='counts' is sparse.")
  }

  # build groups
  gvec <- seu@meta.data[[group.by]]
  gvec <- as.character(gvec)
  keep <- !is.na(gvec)
  if (!all(keep)) {
    if (verbose) message("Dropping ", sum(!keep), " cells with NA in ", group.by)
    X   <- X[, keep, drop = FALSE]
    gvec <- gvec[keep]
  }

  # drop tiny groups
  tab <- table(gvec)
  good_groups <- names(tab)[tab >= min_cells]
  if (length(good_groups) == 0) stop("No groups have >= min_cells.")
  if (length(good_groups) < length(tab) && verbose) {
    message("Dropping ", sum(tab < min_cells), " groups with < ", min_cells, " cells.")
  }
  idx_list <- split(seq_along(gvec), f = gvec)
  idx_list <- idx_list[good_groups]  # keep order stable

  # sparse column-sums per group (genes x groups)
  pb_sum <- do.call(cbind, lapply(idx_list, function(ix)
    Matrix::rowSums(X[, ix, drop = FALSE])
  ))
  colnames(pb_sum) <- good_groups

  # sizes
  n_cells <- vapply(idx_list, length, integer(1))
  lib_sizes_sum <- Matrix::colSums(pb_sum)          # total counts per pseudobulk (sum)
  names(n_cells) <- colnames(pb_sum)

  # adjust if requested
  if (adjust == "none") {
    pb_adj <- NULL
  } else if (adjust == "percell_mean") {
    pb_adj <- Matrix::t(Matrix::t(pb_sum) / n_cells)   # divide each column by its #cells
  } else if (adjust == "percell_cpm") {
    # first average per cell, then CPM using each sample's average-per-cell library size
    pb_mean <- Matrix::t(Matrix::t(pb_sum) / n_cells)
    lib_mean <- Matrix::colSums(pb_mean)
    pb_adj <- Matrix::t(Matrix::t(pb_mean) / pmax(lib_mean, 1)) * 1e6
  }

  out <- list(
    pseudobulk_sum   = pb_sum,           # genes x samples (raw sums)
    adjusted         = pb_adj,           # NULL unless adjust != "none"
    n_cells          = n_cells,          # cells contributing to each sample
    lib_sizes_sum    = lib_sizes_sum,    # library sizes of pseudobulk sums
    groups           = colnames(pb_sum),
    assay            = assay, slot = slot, group.by = group.by, adjust = adjust
  )
  return(out)
}
```

---

## Cohort 1 — Moffitt: loading and quality control

Here we load the Moffitt tumour object and inspect its size. We then run the standard
single-cell **quality control (QC)** checks and visualize them:

- `nFeature_RNA` — how many genes are detected per cell.
- `nCount_RNA` — total transcripts (depth) per cell.
- `percent.mt` — fraction of mitochondrial reads (a marker of stressed/dying cells).

Violin plots and scatter plots help us see where the bulk of "good" cells sit.

```{r moffitt-load-qc, eval=FALSE}
tumor=readRDS(".../Pt_Malignant_plasma_pt1_excluded_Pre_Tx.RDS")
dim(tumor@meta.data) # 41515    58
dim(Assays(tumor,slot = "RNA")) # [1] 31478 41515
Idents(tumor) <- "tumor"  # set contrast as identity
VlnPlot(tumor, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
plot1 <- FeatureScatter(tumor, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(tumor, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2
```

We then **filter out poor-quality cells and likely doublets**. We keep cells with a
reasonable number of genes (not too few = empty/low-quality droplets, not too many =
likely doublets), cap the extreme sequencing depth, and remove cells with high
mitochondrial content.

```{r moffitt-filter, eval=FALSE}
tumor <- subset(
  tumor,
  subset = nFeature_RNA >= 500 &
    nFeature_RNA <= 7500 & # trim extreme high-complexity (often doublets)
    nCount_RNA   <= 75000 & # # trim extreme depth (doublets/aggregates)
    percent.mt   <= 15
)
```

We also load a baseline immune-microenvironment object.

```{r moffitt-immune, eval=FALSE}
immune=readRDS("./baseline_iTME.rds")
dim(immune@meta.data) # 28313   110
dim(Assays(immune,slot = "RNA"))
```

---

## Cohort 2 — Dhodapkar / public: loading and labeling

We load the public Dhodapkar dataset and look at its size and metadata. The key
step here is **defining the clinical outcome label**: patients with progression-free
survival (PFS) over 180 days are called **DR (durable response)**, everyone else
**NDR (non-durable response)**. We also copy the sample identifier into a tidy `sample`
column.

```{r dhodapkar-load, eval=FALSE}
all=readRDS(".../MM_CART_BCD_integrated50pc_Annotated_ClinicalData_infercnv_20231119.rds")
dim(all@meta.data) # 202064    306
head(all@meta.data)
all@meta.data$D270_response=ifelse(all@meta.data$IL18_PFS.group.2 == ">180days","DR","NDR")
all@meta.data$sample=all@meta.data$orig.ident
unique(all@meta.data$IL18_PFS.group)
unique(all@meta.data$IL18_PFS.group.1)
unique(all@meta.data$IL18_PFS.group.2)
dim(Assays(all,slot = "RNA")) # [1]  26298 202064
```

---

## Merging the two cohorts

Now we combine the two datasets into a single object so they can be analyzed together.


```{r merge, eval=FALSE}
## 0) Make sure RNA counts are present & sparse
DefaultAssay(tumor) <- "RNA"
DefaultAssay(all)   <- "RNA"
stopifnot(inherits(GetAssayData(tumor, "RNA", "counts"), "dgCMatrix"))
stopifnot(inherits(GetAssayData(all,   "RNA", "counts"), "dgCMatrix"))
## 2) Ensure unique cell names and add a dataset tag
# tumor <- RenameCells(tumor, add.cell.id = "tumor")
# all   <- RenameCells(all,   add.cell.id = "all")
tumor$dataset <- "moffitt"
all$dataset   <- "public"
## 3) Merge (union of genes; union of metadata columns)
merged <- merge(
  x = tumor, y = all,
  add.cell.ids = c("tumor", "all"),
  project = "merged",
  merge.data = FALSE  # don't try to merge normalized 'data' slots
)
# saveRDS(merged, file = ".../combined_datasets.RDS")
# rm(list=c("all","tumor"))
# merged=readRDS(".../combined_datasets.RDS")
```

A quick sanity check on the merged object and its QC metrics:

```{r merge-inspect, eval=FALSE}
## 4) Inspect
dim(merged)                 # genes x total cells (genes are the union) 4372 243579
ncol(merged@meta.data)      # metadata columns (union; missing become NA)
table(merged$dataset)
Idents(merged) <- "merged"
VlnPlot(merged, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
plot1 <- FeatureScatter(merged, feature1 = "nCount_RNA", feature2 = "percent.mt",group.by = "dataset")
plot2 <- FeatureScatter(merged, feature1 = "nCount_RNA", feature2 = "nFeature_RNA",group.by = "dataset")
plot1 + plot2
```

---

## Pre-processing: normalization, variable genes, scaling, PCA

This is the standard Seurat pre-processing recipe:

1. **Normalize**
2. **Find variable features** 
3. Quickly **inspect** the top variable genes.
4. **Scale** the data and run **PCA** to compress the genes into a handful of summary
   dimensions.

```{r preprocess, eval=FALSE}
merged <- NormalizeData(merged)
merged <- FindVariableFeatures(merged, selection.method = "vst", nfeatures = 2000)

# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(merged), 10)

# plot variable features with and without labels
plot1 <- VariableFeaturePlot(merged)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot1 + plot2

all.genes <- rownames(merged)
merged <- ScaleData(merged, features = all.genes)
merged <- RunPCA(merged, features = VariableFeatures(object = merged))

print(merged[["pca"]], dims = 1:5, nfeatures = 5)
DimPlot(merged, reduction = "pca") + NoLegend()
```

---

## Clustering and UMAP (whole object)

```{r cluster-umap, eval=FALSE}
merged <- FindNeighbors(merged, dims = 1:10)
merged <- FindClusters(merged, resolution = 0.5)

merged <- RunUMAP(merged, dims = 1:10)
DimPlot(merged, reduction = "umap")
DimPlot(merged, reduction = "umap",group.by = "dataset")
```

---

## Selecting the plasma cells

Multiple myeloma is a cancer of plasma cells, so we now **subset to just the plasma
cells**.

```{r select-plasma, eval=FALSE}
CELLTYPE_COL <- "CELL.TYPE"   # <- change to your column
plasma_levels <- c("Plasma","Plasma cell","Malignant plasma")  # adjust
merged@meta.data$CELL.TYPE=merged@meta.data$predicted.celltype.l3
merged@meta.data$CELL.TYPE=ifelse(is.na(merged@meta.data$CELL.TYPE),merged@meta.data$predicted.celltype.l2,merged@meta.data$CELL.TYPE)
seu_plasma <- subset(merged, subset = !!as.name(CELLTYPE_COL) %in% plasma_levels)
# saveRDS(seu_plasma, file = ".../combined_datasets_plasmacells.RDS")
```

---

## Re-processing the plasma cells and removing batch effect (Harmony)

Now that we're working with only plasma cells, we re-run the pipeline on this smaller,
focused object — and importantly, we **correct for the batch difference between
cohorts**.

```{r plasma-reprocess-harmony, eval=FALSE}
# seu_plasma=readRDS(".../combined_datasets_plasmacells.RDS")
DefaultAssay(seu_plasma) <- "RNA"

# keep only what you need right now
keep_assays <- intersect(c("RNA"), Assays(seu_plasma))
seu_plasma  <- DietSeurat(seu_plasma, assays = keep_assays,
                          counts = TRUE, data = TRUE, scale.data = FALSE)  # drop scale.data now
gc()
DefaultAssay(seu_plasma) <- "RNA"
seu_plasma <- NormalizeData(seu_plasma, verbose = FALSE)
seu_plasma <- FindVariableFeatures(seu_plasma, verbose = FALSE)
vf <- VariableFeatures(seu_plasma)
seu_plasma <- ScaleData(seu_plasma, features = vf, verbose = FALSE)

# Run PCA analysis
seu_plasma <- RunPCA(seu_plasma, features = vf)

# Run Harmony to remove the batch effect
seu_plasma <- RunHarmony(seu_plasma, group.by.vars = "dataset")

# Run UMAP analysis
seu_plasma <- RunUMAP(seu_plasma, reduction = "harmony", dims = 1:50)
DimPlot(seu_plasma, reduction = "umap",group.by = "dataset")
```

---

## Differential expression on plasma cells (single-cell level)

With the plasma cells integrated, we set the cell identity to the clinical outcome and
ask: **which genes differ between non-durable (NDR) and durable (DR) responders?** We use
the Wilcoxon test and focus on our gene of interest, ***TNFRSF17* (BCMA)**, the CAR-T
target, visualizing its expression with a violin plot.

```{r plasma-deg, eval=FALSE}
Idents(seu_plasma) <- "D270_response"
de_plasma <- FindMarkers(
  seu_plasma, ident.1 = "NDR", ident.2 = "DR",
  test.use = "wilcox", min.pct = 0.1, logfc.threshold = 0.25
)
head(de_plasma)
de_plasma[rownames(de_plasma) %in% "TNFRSF17",]
VlnPlot(seu_plasma, features = c("TNFRSF17"))
```

---


## Differential expression on the Harmony-integrated plasma cells

We reload the Harmony-corrected object and repeat the DR-vs-NDR comparison, this time
also exporting the full results table and visualising *TNFRSF17* across several views
(violin, feature plot on UMAP, and UMAPs colored by outcome, cohort, and sample).

```{r harmony-deg, eval=FALSE}
seu_plasma=readRDS(".../combined_datasets_plasmacells_post_harmony.RDS")

Idents(seu_plasma) <- "D270_response"
de_plasma <- FindMarkers(
  seu_plasma, ident.1 = "NDR", ident.2 = "DR",
  test.use = "wilcox", min.pct = 0.1, logfc.threshold = 0.25
)
head(de_plasma)
# write.table(de_plasma, ".../SC_DEanalysis_only_plasmacells.txt",
#             col.names = T, row.names = T, quote = F, sep="\t")
de_plasma[rownames(de_plasma) %in% "TNFRSF17",]
VlnPlot(seu_plasma, features = c("TNFRSF17"))
FeaturePlot(seu_plasma, features = c("TNFRSF17"))
DimPlot(seu_plasma, reduction = "umap",group.by = "D270_response")
DimPlot(seu_plasma, reduction = "umap",group.by = "dataset")
DimPlot(seu_plasma, reduction = "umap",group.by = "sample")
```

---

## Pseudobulk: collapsing plasma cells per sample

For the patient-level analysis we now use the `collapse_by_sample()` function defined at
the top. For **each cohort separately**, we subset to plasma cells, then collapse all
those cells into **one average profile per sample** (`adjust = "percell_mean"`). We also
keep a copy of the metadata alongside the collapsed matrix.

```{r pseudobulk-collapse, eval=FALSE}
### Moffitt tumour - Plasma cells
tumor=readRDS(".../Pt_Malignant_plasma_pt1_excluded_Pre_Tx.RDS")
CELLTYPE_COL <- "predicted.celltype.l2"
plasma_levels <- c("Plasma","Plasma cell","Malignant plasma")
tumor2 <- subset(tumor, subset = !!as.name(CELLTYPE_COL) %in% plasma_levels)
tum2.md=as.data.frame(tumor2@meta.data)
tumor2 <- collapse_by_sample(tumor2, group.by = "sample", adjust = "percell_mean")
tumor2 <- append(tumor2, list(tum2.md))
saveRDS(tumor2, ".../tumor_collapsed_counts_only_plasmacells.RDS")

### Dhodapkar tumour - Plasma cells
all=readRDS(".../MM_CART_BCD_integrated50pc_Annotated_ClinicalData_infercnv_20231119.rds")
CELLTYPE_COL <- "predicted.celltype.l3"
plasma_levels <- c("Plasma","Plasma cell","Malignant plasma")
all.2 <- subset(all, subset = !!as.name(CELLTYPE_COL) %in% plasma_levels)
all2.md=as.data.frame(all.2@meta.data)
all.2 <- collapse_by_sample(all.2, group.by = "orig.ident", adjust = "percell_mean")
all.2 <- append(all.2, list(all2.md))
saveRDS(all.2, ".../dhodapkar_collapsed_counts_only_plasmacells.RDS")
```

---

## Building a per-sample expression + metadata table

Next we turn the collapsed matrices into tidy tables that combine a **panel of selected
genes** (immune checkpoint / exhaustion markers plus the CAR-T targets *TNFRSF17*/BCMA
and *GPRC5D*) with the relevant **clinical metadata** for each sample.

### Moffitt

We pull the adjusted (per-cell averaged) expression, keep only our gene panel, transpose
it so samples are rows, attach the clinical columns (time point, MRD, response, sBCMA),
reduce to one row per sample, and **exclude patient 11** (who died within 30 days of
infusion). Finally we merge expression with metadata by sample.

```{r moffitt-table, eval=FALSE}
tumor2=readRDS(".../tumor_collapsed_counts_only_plasmacells.RDS")

moffitt.exp.adj=tumor2$adjusted
dim(moffitt.exp.adj)  # 31478    11
moffitt.exp.adj.sub=moffitt.exp.adj[rownames(moffitt.exp.adj) %in% c("TNFRSF17", "ADORA2A", "BTLA", "CD160", "CD244", "CD27", "CD63",
                                                                     "CD8A", "CD8B", "CHST12", "CSF1", "CSF2", "CST7", "CTLA4", "CTSD",
                                                                     "CTSW", "CXCR6", "ENTPD1", "FAM3C", "FASL", "GZMB", "HAVCR2",
                                                                     "HOPX", "ID2", "IFNG", "IL10", "LAG3", "LAYN", "MIR155HG", "NKG7",
                                                                     "PDCD1", "PRF1", "TIGIT", "TNFRSF9", "TOX", "VSIR", "CCL3", "CCL4",
                                                                     "HAVCR1", "LAIR1", "XCL1", "KLRD1", "KIR3DL1","GPRC5D"),]
dim(moffitt.exp.adj.sub)  # 42 11
moffitt.exp.adj.sub.t=t(moffitt.exp.adj.sub)
moffitt.df=tumor2[[10]]
moffitt.df.1=moffitt.df[,c("sample","time_point","D90_MRD","D270_response","sBCMA")]
moffitt.df.1=unique(moffitt.df.1)  # 11 samples x 5 cols
### Excluding pt 11 because died in 30 days after infusion
moffitt.df.2=moffitt.df.1[moffitt.df.1$sample != "11aCD138",]
moffitt.sub=merge(moffitt.df.2,moffitt.exp.adj.sub.t,by.x="sample",by.y="row.names",all.x=T)
```

### Dhodapkar

The same idea for the public cohort: subset to the gene panel, transpose, re-derive the
DR/NDR label, attach time point, deduplicate to one row per sample, and merge.

```{r dhodapkar-table, eval=FALSE}
all.2=readRDS(".../dhodapkar_collapsed_counts_only_plasmacells.RDS")

all.exp.adj=all.2$adjusted
dim(all.exp.adj)  # 26298    23
all.exp.adj.sub=all.exp.adj[rownames(all.exp.adj) %in% c("TNFRSF17", "ADORA2A", "BTLA", "CD160", "CD244", "CD27", "CD63",
                                                         "CD8A", "CD8B", "CHST12", "CSF1", "CSF2", "CST7", "CTLA4", "CTSD",
                                                         "CTSW", "CXCR6", "ENTPD1", "FAM3C", "FASL", "GZMB", "HAVCR2",
                                                         "HOPX", "ID2", "IFNG", "IL10", "LAG3", "LAYN", "MIR155HG", "NKG7",
                                                         "PDCD1", "PRF1", "TIGIT", "TNFRSF9", "TOX", "VSIR", "CCL3", "CCL4",
                                                         "HAVCR1", "LAIR1", "XCL1", "KLRD1", "KIR3DL1","GPRC5D"),]
dim(all.exp.adj.sub)  # 42 23
all.exp.adj.sub.t=t(all.exp.adj.sub)
all.df=all.2[[10]]
all.df$D270_response=ifelse(all.df$IL18_PFS.group.2 == ">180days","DR","NDR")
all.df$sample=all.df$orig.ident
all.df.1=all.df[,c("sample","time.point","D270_response")]
colnames(all.df.1)[2]="time_point"
all.df.1=unique(all.df.1)  # 23 samples
all.sub=merge(all.df.1,all.exp.adj.sub.t,by.x="sample",by.y="row.names",all.x=T)
```

---

## Combining both cohorts into one table

We keep only the **baseline (pre-treatment)** samples for the response comparison.

```{r combine, eval=FALSE}
df=plyr::rbind.fill(moffitt.sub,all.sub)
dim(df)  # 33 48
### Selecting only the baseline samples
df.bas=df[df$time_point %in% c("PreTx","Pre"),]
dim(df.bas)
```

---

## Testing baseline *TNFRSF17* (BCMA) by response

This is the central result. We use a one-sided **Wilcoxon test** (asking whether
*TNFRSF17* is *higher* in one response group) and draw a boxplot of baseline BCMA
expression in DR vs NDR patients, with individual samples shown as jittered points.


```{r tnfrsf17-test, eval=FALSE}
df.bas=df[df$time_point %in% c("PreTx","Pre"),]
test=wilcox.test(df.bas$TNFRSF17 ~ df.bas$D270_response,alternative = "greater")
boxplot(df.bas$TNFRSF17 ~ df.bas$D270_response,
        las=1,
        outline=T,
        main="TNFRSF17",
        sub=paste("p.value = ",round(test$p.value,4),sep=""),
        xlab="Outcome",
        ylab="TNFRSF17",
        #ylim=c(0,30),
        col=c("dodgerblue","orange"))
stripchart(df.bas$TNFRSF17 ~ df.bas$D270_response,
           method = "jitter",
           pch = 19,
           cex=0.8,
           col = "grey60",
           vertical = TRUE,
           add = TRUE)
```

---

## Exporting the final combined table

Finally, we write the full combined per-sample table (selected genes + metadata, both
cohorts) to a tab-separated file for downstream use and sharing.

```{r export, eval=FALSE}
write.table(df,
            ".../combined_matrix_plasmacells_selected_genes_27MAY26.txt",
            col.names = T,
            row.names = F,
            quote = F,
            sep="\t")
```

---

## Pseudobulk differential expression (DESeq2)

```{r pb-deg-tables, eval=FALSE}
#### Differential expression analysis
moffitt.exp=tumor2$pseudobulk_sum
all.exp=all.2$pseudobulk_sum
### Selecting the genes that are shared by both cohorts
GENES=intersect(rownames(all.exp),rownames(moffitt.exp))

#### Moffitt: shared genes, samples as rows
moffitt.exp=tumor2$pseudobulk_sum
dim(moffitt.exp)            # 31478    11
moffitt.exp=moffitt.exp[rownames(moffitt.exp) %in% GENES,]
moffitt.exp.t=t(moffitt.exp)
head(moffitt.df.1)
### Excluding pt 11 because died in 30 days after infusion
head(moffitt.df.2)
moffitt.deg=merge(moffitt.df.2,moffitt.exp.t,by.x="sample",by.y="row.names",all.x=T)
dim(moffitt.deg)           # 10 25949

#### Dhodapkar: shared genes, samples as rows
all.exp=all.2$pseudobulk_sum
dim(all.exp)               # 26298    23
all.exp=all.exp[rownames(all.exp) %in% GENES,]
all.exp.t=t(all.exp)
head(all.df.1)
all.deg=merge(all.df.1,all.exp.t,by.x="sample",by.y="row.names",all.x=T)
dim(all.deg)               # 23 25947

#### Combine both cohorts and keep baseline samples only
df.deg=plyr::rbind.fill(moffitt.deg,all.deg)
df.deg.bas=df.deg[df.deg$time_point %in% c("PreTx","Pre"),]
dim(df.deg.bas)            # 16
```

```{r pb-deg-input, eval=FALSE}
#### Creating input data
expr=df.deg.bas[,c(1,6:ncol(df.deg.bas))]
rownames(expr)=expr$sample
expr=expr[,-1]
expr=t(expr)

pdata=df.deg.bas[,1:5]
rownames(pdata)=pdata$sample
pdata=pdata[,-1]
```


```{r pb-deg-deseq2, eval=FALSE}
library(edgeR)
library(DESeq2)
dds <- DESeqDataSetFromMatrix(countData = expr,
                              colData = pdata,
                              design = ~ D270_response)

keep <- rowSums(counts(dds) >= 10) >= ceiling(0.2*ncol(dds))
dds  <- dds[keep,]
dds  <- DESeq(dds)
res  <- results(dds, contrast = c("D270_response","NDR","DR"))
```

---

## Gene set enrichment analysis (GSEA)

To move from individual genes to **biological pathways**, we rank every gene by its
DESeq2 test statistic (so genes most up in NDR are at one end, most up in DR at the
other), then run **fgsea** against a collection of gene sets.

First we build the ranked list, handling missing values and infinities.

```{r gsea-ranks, eval=FALSE}
ranks <- res$stat # in order to put CD19 low VS CD19 high
ok <- !is.na(ranks)
ranks <- ranks[ok]
ranks[ranks == Inf] <- 999
names(ranks) <- rownames(res)[ok]
```

We define a custom **CIN70** proliferation/chromosomal-instability signature, and load
two more gene-set collections from file: a modified Hallmark set used for this paper, and
a bone-marrow plasma-cell identity signature.


```{r gsea-genesets, eval=FALSE}
CIN70 <- c("TPX2", "PRC1", "FOXM1", "CDC2", "C20orf24", "TGIF2", "MCM2", "H2AFZ",
           "TOP2A", "PCNA", "UBE2C", "MELK", "TRIP13", "CNAP1", "MCM7", "RNASEH2A",
           "RAD51AP1", "KIF20A", "CDC45L", "MAD2L1", "ESPL1", "CCNB2", "FEN1", "TTK",
           "CCT5", "RFC4", "ATAD2", "ch-TOG", "NUP205", "CDC20", "CKS2", "RRM2", "ELAVL1",
           "CCNB1", "RRM1", "AURKB", "MSH6", "EZH2", "CTPS", "DKC1", "OIP5", "CDCA8",
           "PTTG1", "CEP55", "H2AFX", "CMAS", "BRRN1", "MCM10", "LSM4", "MTB", "ASF1B",
           "ZWINT", "TOPK", "FLJ10036", "CDCA3", "ECT2", "CDC6", "UNG", "MTCH2", "RAD21",
           "ACTL6A", "GPI", "MGC13096", "SFRS2", "HDGF", "NXT1", "NEK2", "DHCR7", "STK6",
           "Aurora-A", "NDUFAB1", "KIAA0286", "KIF4A")

library(fgsea)
library(msigdbr)
# Modified Hallmark collection (paper-specific) read from a .gmt-style table
setlist<-read.table(".../h.all.v6.2.symbols.gmt_CART_paper.txt",sep="\t",fill=T,header=F,as.is=T)
gs<-vector("list",length=nrow(setlist))
names(gs)<-setlist[,1]
setlist<-setlist[,-c(1:2)]
for(i in 1:nrow(setlist)){
  temp<-as.vector(as.matrix(setlist[i,]))
  temp<-temp[temp!=""]
  gs[[i]]<-temp
}

# Bone-marrow plasma-cell identity signature
bm.pc=read.delim(".../HAY_BONE_MARROW_PLASMA_CELL.v2025.1.Hs.gmt",
                 sep="\t",fill=T,header=F,as.is=T)
gs1<-vector("list",length=nrow(bm.pc))
names(gs1)<-bm.pc[,1]
bm.pc<-bm.pc[,-c(1:2)]
for(i in 1:nrow(bm.pc)){
  temp<-as.vector(as.matrix(bm.pc[i,]))   
  temp<-temp[temp!=""]
  gs1[[i]]<-temp
}
```

We combine all gene sets and run fgsea, then keep the pathway, normalized enrichment
score (NES) and adjusted p-value, ordered by significance.

```{r gsea-run, eval=FALSE}
sets  <- c(gs, list(CIN70=CIN70),gs1)
fg <- fgsea(pathways = sets, stats = ranks,nperm = 10000)
NDR=fg[order(fg$padj), c("pathway","NES","padj")]

write.table(as.data.frame(NDR),
            ".../Malignant_sc_plasma_cells_GSEA.txt",
            col.names = T, row.names = F, quote = F, sep="\t")
```

A classic enrichment ("running score") plot for the plasma-cell identity signature,
comparing NDR vs DR:

```{r gsea-enrichplot, eval=FALSE}
plotEnrichment(pathway = sets[["HAY_BONE_MARROW_PLASMA_CELL"]],
               stats = ranks) +
  labs(title = "PLASMA CELL Identity",subtitle = "NDR VS DR") +
  theme_minimal()
```

Finally, a horizontal bar plot of the significant pathways (FDR < 0.1), colored by which
group they're enriched in (red = NDR, blue = DR).

```{r gsea-barplot, eval=FALSE}
sig.pathways=as.data.frame(NDR[NDR$padj < 0.1,])
library(dplyr)
library(ggplot2)

# 1) Prep data
df <- sig.pathways %>%
  dplyr::mutate(
    direction = ifelse(NES > 0, "NDR Enriched", "DR Enriched"),
    label = gsub("^HALLMARK_", "", pathway),
    label = gsub("_", " ", label),
    label = gsub("REACTIVE OXIGEN", "REACTIVE OXYGEN", label, ignore.case = TRUE)
  )

df_plot <- df %>%
  dplyr::arrange(desc(abs(NES))) %>%
  dplyr::mutate(label = factor(label, levels = rev(unique(label[order(NES)]))))

# 2) Plot
p <- ggplot(df_plot, aes(x = label, y = NES, fill = direction)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 0, linewidth = 0.4, colour = "grey40") +
  scale_fill_manual(values = c("NDR Enriched" = "#E66B6B", "DR Enriched" = "#79C4D6")) +
  coord_flip() +
  labs(title = "Malignant PC NDR vs DR at Baseline", x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.title = element_blank(),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )
print(p)
```

---



# Part 2: Immune microenvironment

Part 1 focused on the cancer cells; this part turns to the **immune microenvironment** —
every cell type *except* the malignant plasma cells. The goal is to learn whether the
immune landscape at baseline (which cell types are present, in what proportions, and in
what activation or exhaustion state) relates to durable response. The workflow mirrors
Part 1: select the relevant cells, integrate the two cohorts with Harmony, then compare
DR vs NDR both within specific cell types and across overall cell composition.

---

## Loading the public data and a marker check

We reload the public dataset, re-create the response label, and pull raw counts for a
panel of immune/exhaustion markers as a quick sanity check.

```{r immune-load, eval=FALSE}
all=readRDS(".../MM_CART_BCD_integrated50pc_Annotated_ClinicalData_infercnv_20231119.rds")
all@meta.data$D270_response=ifelse(all@meta.data$IL18_PFS.group.2 == ">180days","DR","NDR")
all@meta.data$sample=all@meta.data$orig.ident

orig.expr=all@assays$RNA@counts[rownames(all@assays$RNA@counts) %in% c("TNFRSF17", "ADORA2A", "BTLA",
                                                                        "CD160", "CD244", "CD27", "CD63","CD8A", "CD8B",
                                                                        "CHST12", "CSF1", "CSF2", "CST7", "CTLA4", "CTSD",
                                                                        "CTSW", "CXCR6", "ENTPD1", "FAM3C", "FASL", "GZMB", "HAVCR2",
                                                                        "HOPX", "ID2", "IFNG", "IL10", "LAG3", "LAYN", "MIR155HG", "NKG7",
                                                                        "PDCD1", "PRF1", "TIGIT", "TNFRSF9", "TOX", "VSIR", "CCL3", "CCL4",
                                                                        "HAVCR1", "LAIR1", "XCL1", "KLRD1", "KIR3DL1",
                                                                        "PTPN12","HLA-DRB1"),]
```

## Selecting immune cells and merging the cohorts

Here we keep only the **non-plasma, non-doublet** cell types from each cohort (the
inverse of what we did earlier), tag each by dataset, and merge the Moffitt immune object
with the public immune cells into one combined object.

```{r immune-select-merge, eval=FALSE}
## Make sure RNA counts are present & sparse
DefaultAssay(immune) <- "RNA"
DefaultAssay(all)   <- "RNA"
stopifnot(inherits(GetAssayData(immune, "RNA", "counts"), "dgCMatrix"))
stopifnot(inherits(GetAssayData(all,   "RNA", "counts"), "dgCMatrix"))
immune$dataset <- "moffitt"
all$dataset   <- "public"

## Keep every cell type EXCEPT plasma cells and doublets
CELLTYPE_COL <- "predicted.celltype.l3"
plasma_levels <- setdiff(unique(c(unique(all@meta.data$predicted.celltype.l3),unique(immune@meta.data$predicted.celltype.l2))),
                         c("Plasma","Plasma cell","Malignant plasma"))
plasma_levels=plasma_levels[plasma_levels != "Doublet"]
all@meta.data$CELL.TYPE=all@meta.data$predicted.celltype.l2
immune@meta.data$CELL.TYPE=immune@meta.data$predicted.celltype.l2
all.imm <- subset(all, subset = !!as.name(CELLTYPE_COL) %in% plasma_levels)
dim(all.imm)
rm(list = c("all"))

## Merge the two immune compartments
merged <- merge(
  x = immune, y = all.imm,
  add.cell.ids = c("immune", "all"),
  project = "merged",
  merge.data = FALSE
)
dim(merged)
table(merged$dataset)
Idents(merged) <- "merged"
# unify the mitochondrial-percent column name across cohorts
merged@meta.data$percent.mt=ifelse(is.na(merged@meta.data$percent.mt),merged@meta.data$percent_mito,merged@meta.data$percent.mt)
VlnPlot(merged, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)

# unified cell-type label, finer where available
merged@meta.data$CELL.TYPE=merged@meta.data$predicted.celltype.l3
merged@meta.data$CELL.TYPE=ifelse(is.na(merged@meta.data$CELL.TYPE),merged@meta.data$predicted.celltype.l2,merged@meta.data$CELL.TYPE)
rm(list = c("all.imm","immune"))
```

## Pre-processing, clustering and UMAP

The same standard pipeline as the plasma-cell analysis: normalize, find variable genes,
scale, PCA, cluster, and UMAP. The result is saved so this heavy step can be reloaded
later.

```{r immune-preprocess, eval=FALSE}
merged <- NormalizeData(merged)
merged <- FindVariableFeatures(merged, selection.method = "vst", nfeatures = 2000)
top10 <- head(VariableFeatures(merged), 10)

all.genes <- rownames(merged)
merged <- ScaleData(merged, features = all.genes)
merged <- RunPCA(merged, features = VariableFeatures(object = merged))
print(merged[["pca"]], dims = 1:5, nfeatures = 5)
DimPlot(merged, reduction = "pca") + NoLegend()

merged <- FindNeighbors(merged, dims = 1:10)
merged <- FindClusters(merged, resolution = 0.5)
merged <- RunUMAP(merged, dims = 1:10)
# saveRDS(merged, ".../combined_datasets_immune_cells.RDS")

merged=readRDS(".../combined_datasets_immune_cells.RDS")
DimPlot(merged, reduction = "umap")
DimPlot(merged, reduction = "umap",group.by = "dataset")
DimPlot(merged, reduction = "umap",group.by = "CELL.TYPE")
```

## Batch correction with Harmony


```{r immune-harmony, eval=FALSE}
merged=readRDS(".../combined_datasets_immune_cells.RDS")
DefaultAssay(merged) <- "RNA"

keep_assays <- intersect(c("RNA"), Assays(merged))
merged  <- DietSeurat(merged, assays = keep_assays, data = TRUE, scale.data = FALSE)
gc()
DefaultAssay(merged) <- "RNA"
merged <- NormalizeData(merged, verbose = FALSE)
merged <- FindVariableFeatures(merged, verbose = FALSE)
vf <- VariableFeatures(merged)
merged <- ScaleData(merged, features = vf, verbose = FALSE)

merged <- RunPCA(merged, features = vf)
merged <- RunHarmony(merged, group.by.vars = "dataset")
merged <- RunUMAP(merged, reduction = "harmony", dims = 1:50)

DimPlot(merged, reduction = "umap",group.by = "dataset",pt.size = 0.01)
DimPlot(merged, reduction = "umap",group.by = "CELL.TYPE",pt.size = 0.01)
DimPlot(merged, reduction = "umap",group.by = "predicted.celltype.l1",pt.size = 0.01)
DimPlot(merged, reduction = "umap",group.by = "sample",pt.size = 0.01)
```

## Restricting to baseline samples

We keep only **baseline (pre-treatment)** samples.

```{r immune-baseline, eval=FALSE}
merged@meta.data$time.point=merged@meta.data$time_point
library(stringr)
merged@meta.data$time.point=ifelse(is.na(merged@meta.data$time.point),
                     str_split_fixed(merged@meta.data$sample,"_",2)[,2],merged@meta.data$time.point)

baseline <- subset(merged, subset = time.point %in% c("pre","PreTx"))

# rebuild graph/UMAP in the same reduction used before
reduc <- if ("harmony" %in% Reductions(baseline)) "harmony" else "pca"
baseline <- FindNeighbors(baseline, reduction = reduc, dims = 1:30)
baseline <- FindClusters(baseline, resolution = 0.5)
baseline <- RunUMAP(baseline, reduction = reduc, dims = 1:30)

DimPlot(baseline, group.by = "predicted.celltype.l1")
DimPlot(baseline, reduction = "umap",group.by = "dataset",pt.size = 0.01)
DimPlot(baseline, reduction = "umap",group.by = "CELL.TYPE",pt.size = 0.01)
DimPlot(baseline, reduction = "umap",group.by = "sample",pt.size = 0.01)
# saveRDS(merged, ".../combined_datasets_immune_cells_postHarmony.RDS")
```

## Differential expression within CD4 T cells

Here we test DR vs NDR **within a specific cell type** — CD4 T cells — by setting the
identity to the cell-type label and using `subset.ident` to restrict the comparison. We
then keep the exhaustion-related genes of interest.

```{r immune-deg, eval=FALSE}
Idents(baseline) <- "predicted.celltype.l1"   # identities are the cell types
de_plasma <- FindMarkers(
  baseline,
  ident.1 = "NDR", ident.2 = "DR",
  group.by = "D270_response",   # compare DR vs NDR
  subset.ident = c("CD4 T"),    # only within CD4 T cells
  test.use = "wilcox", min.pct = 0.01, logfc.threshold = 0.1
)

new=de_plasma[rownames(de_plasma) %in% c("CD160", "CST7", "GZMB", "HAVCR2",
                                       "LAG3", "PDCD1","TIGIT", "KLRD1",
                                       "PTPN12","HLA-DRB1", "TOX"),]
new$gene=rownames(new)
de_plasma$gene=rownames(de_plasma)
```

## Exhaustion dot plot (CD4 T)

A dot plot summarizing the exhaustion genes: dot **size** = percent of cells expressing
the gene, dot **color** = log2 fold-change direction (red toward NDR, blue toward DR),
and **stars** marking significance.

```{r immune-dotplot, eval=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)

pstars <- function(p) {
  ifelse(p < 1e-4, "****",
         ifelse(p < 1e-3, "***",
                ifelse(p < 1e-2, "**",
                       ifelse(p < 0.05, "*", ""))))
}

df_long <- new %>%
  transmute(
    gene,
    FDR = p_val_adj,
    lfc = avg_log2FC,
    NDR_pct = pct.1 * 100,
    DR_pct  = pct.2 * 100
  ) %>%
  pivot_longer(c(NDR_pct, DR_pct), names_to = "group", values_to = "pct") %>%
  mutate(
    D270_response = ifelse(group == "NDR_pct", "NDR", "DR"),
    color_val = ifelse(D270_response == "NDR", lfc, -lfc),
    stars = pstars(FDR),
    p_label = sprintf("FDR=%.2g", FDR)
  )

gene_order <- rev(df_long %>% distinct(gene) %>% pull(gene))
df_long$gene <- factor(df_long$gene, levels = gene_order)

ggplot(df_long, aes(x = D270_response, y = gene)) +
  geom_point(aes(size = pct, fill = color_val),
             shape = 21, colour = "grey25", stroke = 0.25) +
  geom_text(aes(label = stars), vjust = -0.9, size = 3.5, lineheight = 0.9, na.rm = TRUE) +
  scale_size_continuous("Percent Expressed", range = c(1.8, 7),
                        breaks = c(25, 50, 75), limits = c(0, 100)) +
  scale_fill_gradient2("log2(FC)",
                       low = "#2C7FB8", mid = "white", high = "#D7301F",
                       midpoint = 0, limits = c(-4, 4)) +
  labs(title = "Genes related to exhaustion - CD4 T", x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.y = element_blank(),
    legend.position = "right",
    plot.title = element_text(face = "bold", hjust = 0.5)
  )
```

## Per-sample immune expression table


```{r immune-expr-table, eval=FALSE}
expr=merged@assays$RNA@scale.data[rownames(merged@assays$RNA@scale.data) %in% c("TNFRSF17", "ADORA2A", "BTLA",
                                                                        "CD160", "CD244", "CD27", "CD63","CD8A", "CD8B",
                                                                        "CHST12", "CSF1", "CSF2", "CST7", "CTLA4", "CTSD",
                                                                        "CTSW", "CXCR6", "ENTPD1", "FAM3C", "FASL", "GZMB", "HAVCR2",
                                                                        "HOPX", "ID2", "IFNG", "IL10", "LAG3", "LAYN", "MIR155HG", "NKG7",
                                                                        "PDCD1", "PRF1", "TIGIT", "TNFRSF9", "TOX", "VSIR", "CCL3", "CCL4",
                                                                        "HAVCR1", "LAIR1", "XCL1", "KLRD1", "KIR3DL1",
                                                                        "PTPN12","HLA-DRB1"),]
expr=as.data.frame(expr)
expr.t=t(expr)

pdata=merged@meta.data
pdata=as.data.frame(pdata)
pdata=pdata[,1:112]

df=merge(pdata,expr.t,by="row.names")
df$time.point=df$time_point
library(stringr)
df$time.point=ifelse(is.na(df$time.point),
                     str_split_fixed(df$sample,"_",2)[,2],df$time.point)

write.table(df, ".../combined_matrix_immunecells_selected_genes_scaled_data.txt",
            col.names = T, row.names = F, quote = F, sep="\t")
```

## Cell-type composition by response and cohort

This part counts how many cells of each type each sample has, then tests whether **cell
composition differs by outcome (DR vs NDR) and by cohort**. We reload a pre-built
per-sample count table, restrict to baseline, exclude patient 11, and run Wilcoxon tests
with boxplots — starting with the plasma-cell counts.

```{r immune-composition, eval=FALSE}
tab.cells.1=read.delim(".../number_of_cells_per_each_sample.txt", stringsAsFactors = F)
tab.cells.1.bas=tab.cells.1[tab.cells.1$time.point == "pre",]
tab.cells.1.bas=tab.cells.1.bas[tab.cells.1.bas$sample != "11a",]

wilcox.test(tab.cells.1.bas$Plasma ~ tab.cells.1.bas$D270_response)
boxplot(tab.cells.1.bas$Plasma ~ tab.cells.1.bas$D270_response)
wilcox.test(tab.cells.1.bas$Plasma ~ tab.cells.1.bas$dataset)
boxplot(tab.cells.1.bas$Plasma ~ tab.cells.1.bas$dataset)
```

We then loop over each cell-type column, producing a boxplot + Wilcoxon test for **every
cell type vs cohort** and **vs outcome**, saving each as a PDF. The loop is run twice on
the raw counts (one set of PDFs per grouping variable).

```{r immune-composition-loops, eval=FALSE}
# Enrichment of each cell type by DATASET (raw counts)
for(i in 5:15){
  pdf(paste(".../Enrichment_based_on_dataset/", colnames(tab.cells.1.bas)[i],"_VS_Dataset.pdf",sep=""),width = 5,height = 4)
  test<-wilcox.test(tab.cells.1.bas[,i] ~ tab.cells.1.bas$dataset)
  boxplot(tab.cells.1.bas[,i] ~ tab.cells.1.bas$dataset,
          las=1, outline=T, main=colnames(tab.cells.1.bas)[i],
          sub=paste("p.value = ",round(test$p.value,4),sep=""),
          xlab="Dataset", ylab=paste("# of ",colnames(tab.cells.1.bas)[i],sep=""),
          col=c("dodgerblue","orange"))
  stripchart(tab.cells.1.bas[,i] ~ tab.cells.1.bas$dataset,
             method = "jitter", pch = 19, cex=0.8, col = "grey60", vertical = TRUE, add = TRUE)
  dev.off()
}

# Enrichment of each cell type by OUTCOME (raw counts)
for(i in 5:15){
  pdf(paste(".../Enrichment_based_on_outcome/", colnames(tab.cells.1.bas)[i],"_VS_Outcome.pdf",sep=""),width = 5,height = 4)
  test<-wilcox.test(tab.cells.1.bas[,i] ~ tab.cells.1.bas$D270_response)
  boxplot(tab.cells.1.bas[,i] ~ tab.cells.1.bas$D270_response,
          las=1, outline=T, main=colnames(tab.cells.1.bas)[i],
          sub=paste("p.value = ",round(test$p.value,4),sep=""),
          xlab="Outcome", ylab=paste("# of ",colnames(tab.cells.1.bas)[i],sep=""),
          col=c("dodgerblue","orange"))
  stripchart(tab.cells.1.bas[,i] ~ tab.cells.1.bas$D270_response,
             method = "jitter", pch = 19, cex=0.8, col = "grey60", vertical = TRUE, add = TRUE)
  dev.off()
}
```

Counts can be misleading if samples have very different total cell numbers, so we convert
each cell type into a **percentage of total immune cells** and repeat the same two loops
on those proportions.

```{r immune-composition-pct, eval=FALSE}
# convert counts to percentage of total immune cells
for(i in 5:15){
  tab.cells.1.bas[,ncol(tab.cells.1.bas) + 1]=tab.cells.1.bas[,i] / tab.cells.1.bas$tot.immune
  colnames(tab.cells.1.bas)[ncol(tab.cells.1.bas)]=paste("perc.",colnames(tab.cells.1.bas)[i],sep="")
}

# Percentages by DATASET
for(i in 18:27){
  pdf(paste(".../Enrichment_based_on_dataset/Percentage_", colnames(tab.cells.1.bas)[i],"_VS_Dataset.pdf",sep=""),width = 5,height = 4)
  test<-wilcox.test(tab.cells.1.bas[,i] ~ tab.cells.1.bas$dataset)
  boxplot(tab.cells.1.bas[,i] ~ tab.cells.1.bas$dataset,
          las=1, outline=T, main=colnames(tab.cells.1.bas)[i],
          sub=paste("p.value = ",round(test$p.value,4),sep=""),
          xlab="Dataset", ylab=paste("% of ",colnames(tab.cells.1.bas)[i],sep=""),
          col=c("dodgerblue","orange"))
  stripchart(tab.cells.1.bas[,i] ~ tab.cells.1.bas$dataset,
             method = "jitter", pch = 19, cex=0.8, col = "grey60", vertical = TRUE, add = TRUE)
  dev.off()
}

# Percentages by OUTCOME
for(i in 18:27){
  pdf(paste(".../Enrichment_based_on_outcome/Percentage_", colnames(tab.cells.1.bas)[i],"_VS_Outcome.pdf",sep=""),width = 5,height = 4)
  test<-wilcox.test(tab.cells.1.bas[,i] ~ tab.cells.1.bas$D270_response)
  boxplot(tab.cells.1.bas[,i] ~ tab.cells.1.bas$D270_response,
          las=1, outline=T, main=colnames(tab.cells.1.bas)[i],
          sub=paste("p.value = ",round(test$p.value,4),sep=""),
          xlab="Outcome", ylab=paste("% of ",colnames(tab.cells.1.bas)[i],sep=""),
          col=c("dodgerblue","orange"))
  stripchart(tab.cells.1.bas[,i] ~ tab.cells.1.bas$D270_response,
             method = "jitter", pch = 19, cex=0.8, col = "grey60", vertical = TRUE, add = TRUE)
  dev.off()
}
```

Finally, we combine CD4 and CD8 T cells into a single **total T-cell percentage** and
test whether it differs between DR and NDR, plus a fresh per-sample cell-type table.

```{r immune-tcell-pct, eval=FALSE}
tab.cells.1.bas$perc.T.cells=(tab.cells.1.bas$CD4.T + tab.cells.1.bas$CD8.T) / tab.cells.1.bas$tot.immune
test<-wilcox.test(tab.cells.1.bas$perc.T.cells ~ tab.cells.1.bas$D270_response)
boxplot(tab.cells.1.bas$perc.T.cells ~ tab.cells.1.bas$D270_response,
        las=1, outline=T, main="Percentage of T cells",
        sub=paste("p.value = ",round(test$p.value,4),sep=""),
        xlab="Outcome", ylab=paste("% of ","T cells",sep=""),
        col=c("dodgerblue","orange"))
stripchart(tab.cells.1.bas$perc.T.cells ~ tab.cells.1.bas$D270_response,
           method = "jitter", pch = 19, cex=0.8, col = "grey60", vertical = TRUE, add = TRUE)

tab.cells1=as.data.frame.matrix(table(df$sample,df$CELL.TYPE))
tab.cells1$tot=rowSums(tab.cells1)
```

---

# Session information

Including `sessionInfo()` records the exact package versions used, which reviewers and
readers appreciate for reproducibility.

```{r session-info, eval=TRUE}
sessionInfo()
```
